\documentclass[sigplan,screen,review]{acmart}

% Required Packages
\usepackage{amsmath}      % For math symbols like \neg, \Rightarrow
\usepackage{diagbox}      % For diagonally split table cells
\usepackage[table]{xcolor}% For cell coloring (\textcolor)
\usepackage{graphicx}     % For rotating text (\rotatebox) - used for
% table headers

% Optional: Adjust page margins if needed
% \usepackage[margin=1in]{geometry}

% --- Custom Commands for Table Elements (optional but recommended) ---
% T for True (Green)
\newcommand{\tableT}{\textcolor{green}{T}}
% F for False (Red)
\newcommand{\tableF}{\textcolor{red}{F}}
% N/A or self-comparison cell (using a dash here, adjust as needed)
\newcommand{\tableNA}{-}
% Command for rotated table headers (adjust angle and style if needed)
\newcommand{\rotatedHeader}[1]{\rotatebox{90}{\textit{#1}}}
% ---------------------------------------------------------------------

\acmJournal{PACMPL}
\acmVolume{7}
\acmNumber{POPL}
\acmArticle{1}
\acmYear{2025}
\acmMonth{5}
\acmDOI{}

\setcopyright{acmlicensed}
\copyrightyear{2025}

\AtBeginDocument{%
  \providecommand\BibTeX{{%
Bib\TeX}}}

\begin{document}

\title{The Name of the Title is Hope}

\author{Yifei Sun}
\email{ysun@ccs.neu.edu}
\orcid{0000-0002-1591-7458}

\author{Ji-Yong Shin}
\email{j.shin@northeastern.edu}
\orcid{0000-0002-1595-4849}

\affiliation{%
  \institution{Northeastern University}
  \city{Boston}
  \state{Massachusetts}
  \country{USA}
}

\renewcommand{\shortauthors}{Sun et al.}

\begin{abstract}
  Your abstract goes here. This should be a concise summary of your work.
\end{abstract}

% CCS Concepts: https://dl.acm.org/ccs/ccs.cfm
\begin{CCSXML}
  <ccs2012>
  <concept>
  <concept_id>10011007.10011006.10011008</concept_id>
  <concept_desc>Software and its engineering~Software creation and
  management</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Software creation and management}

\keywords{Your keywords here, separated by commas}

\maketitle

\input{introduction}
\input{terminology}
\input{consistency-semantics}
\input{semantic-comparison-and-composition}
\input{modeling-real-world-applications}

\section{Background}
Background section content.

\section{Method}
Method section content.

\section{Evaluation}
Evaluation section content.

\section{Related Work}
Related work section content.

\section{Conclusion}
Conclusion section content.

\bibliographystyle{acmart}
\bibliography{references}

\end{document}
